Staged Abstract Interpreters

### 2019-09-07 paper/10_intro.tex
typos

Roughly at the same time when abstract interpretation was proposed in the
1970s, \citet{futamura1971partial} observed a close
connection between interpreters and compilers through a hierarchy of
specializations, since then known as Futamura Projections. The first Futamura
projection describes that specializing an interpreter to the input program
removes interpretative overhead and yields an equivalent executable
that runs faster. 
In recent years, the idea of Futamura projection has regained more attentions
and successfully applied to many real-world problems, examples including
query compilers \cite{DBLP:conf/sigmod/TahboubER18}, PyPy \cite{Bolz:2009:TMP:1565824.1565827},
Truffle \cite{Marr:2015:TVP:2814270.2814275}, etc.
The challenge here: can we apply the same idea to a general abstract interpreter
to make it run faster? The solution is not obvious, as we need to specialize
a nonstandard semantics that extensively uses nondeterminism, operates on
abstract domains, and computes fixed-points iteratively over a cache.

@@

Although the four artifacts may look dissimilar at first glance, they in
fact are all firmly rooted in the concrete semantics of the language.  This
observation provides a way to abstract over the interpreter and achieve the
flexibility of reinterpreting the shared interpreter.  The generic interpreter
returns a value of monadic type, which can be varied by different semantics.
The value domains of the interpreter and the effects such as state and nondeterminism
can all be wrapped into this monadic type.  The binding-time abstraction is
represented by a higher-kinded type, and injected into the monadic type.  The
binding-times control whether the interpreter produces values directly or
generates code.

>>>>

Roughly at the same time when abstract interpretation was proposed in the
1970s, \citet{futamura1971partial} observed a close
connection between interpreters and compilers through a hierarchy of
specializations, since then known as Futamura Projections. The first Futamura
projection describes that specializing an interpreter to the input program
removes interpretative overhead and yields an equivalent executable
that runs faster. 
In recent years, the idea of Futamura projections has gained more attentions
and has been successfully applied to many real-world problems, examples including
query compilers \cite{DBLP:conf/sigmod/TahboubER18}, PyPy \cite{Bolz:2009:TMP:1565824.1565827},
Truffle \cite{Marr:2015:TVP:2814270.2814275}, etc.
The challenge here: can we apply the same idea to a general abstract interpreter
to make it run faster? The solution is not obvious, as we need to specialize
a nonstandard semantics that extensively uses nondeterminism, operates on
abstract domains, and computes fixed-points iteratively over caches.

@@

Although the four artifacts may look dissimilar at first glance, they in
fact are all firmly rooted in the concrete semantics.  This
observation provides a way to abstract over the interpreter and achieve the
flexibility of reinterpreting the shared interpreter.  The generic interpreter
returns a value of monadic type, which can be varied by different semantics.
The value domains of the interpreter and the effects such as state and nondeterminism
can all be wrapped into this monadic type.  The binding-time abstraction is
represented by a higher-kinded type, and injected into the monadic type.  The
binding-times control whether the interpreter produces values directly or
generates code.

### 2019-09-07 paper/50_stageabs.tex
typos

In the previous sections, we have seen an unstaged abstract interpreter and a
staged concrete interpreter. Now we begin describing the implementation of
their confluence -- a staged abstract interpreter.  Unsurprisingly, the staged
abstract interpreter in this section uses the same abstract semantics from the
unstaged version in Section~\ref{unstaged_abs}.  The approach we refactor
the unstaged one to the staged abstract interpreter is modular, and does not
sacrifice soundness or precision. The designer of the analyzer therefore has no
need to rewrite the analysis. We first present the staged lattices and staged
monads, and then discuss the staged version of primitive operations, especially
@close@, @ap_clo@ and @fix@. At last, we discuss several optimizations.

@@

\paragraph{Staged Monads for Abstract Interpretation}
We use the same monad stack structure from the unstaged abstract interpreter,
but replacing \textit{all} the monad transformers to their staged version. The
following code snippet shows this change. We manually fuse the three inner
transformers (@IdM@ omitted) into a single monad @RepSetReaderStateM[A, B, C]@,
where @A@ is the type parameter for the reader effects and @B@ for the state effects.
In our monad stack scheme, they are both instantiated with the type @Cache@. 
Similar to the unstaged version, the grounded result type is a pair of two
staged value: @Rep[Set[(Value, Store)]]@ and @Rep[Cache]@.
\begin{lstlisting}
  trait StagedAbstractSemantics extends AbstractComponents {
    type R[T] = Rep[T]
    type AnsM[T] = RepReaderT[RepStateT[RepSetReaderStateM[Cache, Cache, ?], Store, ?], Env, T]
    type Result = (Rep[Set[(Value, Store)]], Rep[Cache])
    ...
  }
\end{lstlisting}

>>>>

In the previous sections, we have seen an unstaged abstract interpreter and a
staged concrete interpreter. Now we begin describing the implementation of
their confluence -- a staged abstract interpreter.  Unsurprisingly, the staged
abstract interpreter in this section uses the same abstract semantics from the
unstaged version in Section~\ref{unstaged_abs}.  The approach we use to refactor
the unstaged one to the staged abstract interpreter is modular, and does not
sacrifice soundness or precision. The designer of the analyzer therefore has no
need to rewrite the analysis. We first present the staged lattices and staged
monads, and then discuss the staged version of primitive operations, especially
@close@, @ap_clo@ and @fix@. In the end, we discuss several optimizations.

@@

\paragraph{Staged Monads for Abstract Interpretation}
We use the same monad stack structure from the unstaged abstract interpreter,
but replacing \textit{all} the monad transformers with their staged versions. The
following code snippet shows this change. We manually fuse the three inner
transformers (@IdM@ omitted) into a single monad @RepSetReaderStateM[A, B, C]@,
where @A@ is the type parameter for the reader effects and @B@ for the state effects.
In our monad stack scheme, they are both instantiated with the type @Cache@. 
Similar to the unstaged version, the grounded result type is a pair of two
staged value: @Rep[Set[(Value, Store)]]@ and @Rep[Cache]@.
\begin{lstlisting}
  trait StagedAbstractSemantics extends AbstractComponents {
    type R[T] = Rep[T]
    type AnsM[T] = RepReaderT[RepStateT[RepSetReaderStateM[Cache, Cache, ?], Store, ?], Env, T]
    type Result = (Rep[Set[(Value, Store)]], Rep[Cache])
    ...
  }
\end{lstlisting}
